---
title: "Sensitivity analyses"
format: html
---

The list of packages required for reproducing the analyses on this script are listed in @lst-setup-sensitivity-analyses: 

```{r setup-prcc}
#| lst-label: lst-setup-sensitivity-analyses
#| lst-cap: Packages required for computing sensitivity analyses. 
# remotes::install_github("jpahle/CoRC")
library(CoRC)

# tidyverse packages
library(dplyr)
library(ggplot2)

# to generate the simulations 
library(lhs) # Latin Hypercube sampling
library(parallel) # parallel computing
library(ppcor) # Compute partial correlations

# Table reporting
library(flextable)
```

## Evaluate the sensitivity of immune cells to parameter constants

### Generate the distributions of nine critical parameters 

We begin to extract the original constants of the 9 parameters assumed to have the strongest impact on the populations of T cell subsets, using `CoRC::getParameters`, reported in @tbl-healthy-parameters:  

```{r}
#| label: tbl-healthy-parameters
#| tbl-cap: The 9 reaction parameters evaluated for **sensitivity analyses**. 
## Retrieve parameters we want to evaluate the versatility ----
healthy_model <- loadModel("../../models/team_2016_final_model_lo2016_2025_05_13.cps")
# S2 to be changed
parameters_healthy <- getParameters(model = healthy_model,
                                  key = c("(Diff of M0 to M1).delta_m_cit",
                                          "(Diff of M1 to M2).sigma",
                                          "(Induction of T1 from M).s12",
                                          "(Proliferation of T1).s2",
                                          "(Induction of T2).s4",
                                          "(Induction of T17).s21",
                                          "(Induction of T17).s6",
                                          "(Induction of Tr).sb",
                                          "(Induction of Tr).s10")) |> 
  dplyr::select(-mapping)

flextable(parameters_healthy) |> 
  add_footer_row(values = "key is direct ID of the parameter in the model, reaction describes the biological mechanism associated with the value of the parameter, and value returns the constant value assumed for healthy individuals.", 
                 colwidths = 4) |> 
  bold(part = "header") 
```


In @lo2016po, they chose to perform **Latin Hypercube Sampling (LHS)** against **Standard Random Sampling**. The differences between these two randomisation approaches are further detailed in @tip-LHS, but briefly, the major advantage of LHS lies in its better coverage of the whole parameter sampling space, especially with a low number of observations (which is usually the case in bootstrap simulations due to their high computing cost). 

::: {#tip-LHS .callout-tip collapse="true" title="Standard versus Latin Hypercube Sampling (LHS)"}

Random sampling and Latin Hypercube Sampling differ in how they distribute the sample points within the parameter space:

---

**Random Sampling** for a $d-$ random vector in : 

$$
x_i^{(j)} \sim \mathcal{U}\left(x_{\min_j}^{(j)}, x_{\max_j}^{(j)}\right), \quad j = 1, 2, \dots, d, \quad i = 1, 2, \dots, N
$$ {#eq-random-sampling}

where $x_i^{(j)}$ is the $i$-th sample drawn for the $j$-th dimension, $d=9$ (number of parameters evaluated for sensitivity), $N=5000$ the total number of parameter distributions simulated, and $p_j$ the probability distribution^[Only the uniform sampling, $U_j \sim \mathcal{U}(0,1)$, is provided with the `LHS`package. We detailed in @tip-ITST how to simulate any kind of probability distribution from the standard uniform distribution.].
Since each sample is drawn completely at random, it's quite likely that some regions might be over-sampled while others may have gaps, with a low number of observations drawn (good coverage of the whole sampling space is only guaranteed asymptotically).

--- 

In **Latin Hypercube Sampling (LHS)**,  the sampling space is first stratified into $N$ equally probable intervals for each variable $x^{(j)}$; an independent permutation is subsequently applied across dimensions to avoid correlation. Finally, each sample is drawn from of the $N-$ defined intervals.
LHS then guarantees a more evenly distributed coverage.

---

To conclude, LHS provides a **better Coverage of the Input Space**(notably reduces the probability of clustering or missing critical regions), **Improved Convergence and Efficiency** by achieving a more accurate representation of the underlying distribution and leading to lower variance in estimated statistics with fewer number of available observations (see @fig-LHS for further illustration). Yet, only standard random sampling is completely independent and unbiased. 


![Randomly sampled variables vs. Latin hypercube samples, from [@rustell](https://www.researchgate.net/figure/Randomly-sampled-variables-vs-Latin-hypercube-samples_fig3_322386503). With a low number of observations, the **clustering pattern** inherent to standard random sampling is clearly showcased.](../assets/img/Randomly-sampled-variables-vs-Latin-hypercube-samples.png){#fig-LHS}

:::

The code snippet below generates the $N=5000$ bootstrap distribution of the 9 evaluated parameters for robustness, using **Latin Hypercube Sampling**. LHS indeed guarantees better coverage of the *sampling space*.^[Note however that both standard random sampling and LHS assume independence between the parameters values, a strong and quite unlikely biological premise.].

```{r}
#| label: lhs-generation
n_samples <- 5000
# Lower and upper bounds are computed within a ±20% range centered around the assumed parameter constant.
parameters_sensitivity <- parameters_healthy$value
param_ranges <- tibble::tibble(parameters_name = parameters_healthy$key, 
                               LB= parameters_sensitivity *0.8,
                               UB = parameters_sensitivity *1.2)
set.seed(20) 
bootstrap_parameters <- randomLHS(n_samples, length(parameters_sensitivity))

for (i in seq_along(parameters_sensitivity)) {
  bootstrap_parameters[,i] <- qunif(bootstrap_parameters[,i], 
                                    param_ranges$LB[i],
                                    param_ranges$UB[i])
}
colnames(bootstrap_parameters) <- parameters_healthy$name
apply(bootstrap_parameters,2, quantile)
```


```r
n_samples <- 5000 # <1>

parameters_sensitivity <- parameters_healthy$value
param_ranges <- tibble::tibble(parameters_name = parameters_healthy$key, # <2>
                               LB= parameters_sensitivity *0.8,          # <2>
                               UB = parameters_sensitivity *1.2)         # <2>


set.seed(20) # <3>
bootstrap_parameters <- randomLHS(n_samples, length(parameters_sensitivity)) # <4>

# 
for (i in seq_along(parameters_sensitivity)) { # <5>
  bootstrap_parameters[,i] <- qunif(bootstrap_parameters[,i], # <5>
                                    param_ranges$LB[i],       # <5>
                                    param_ranges$UB[i])       # <5>
}                                                             # <5>
colnames(bootstrap_parameters) <- parameters_healthy$name
apply(bootstrap_parameters,2, quantile)                       # <6>
```
1. Determine the number of observations, equivalently the number of generated bootstrap distributions (here, $N=5000$).
2. Lower and upper bounds are computed within a $\pm 20\%$ range, centred around the parameter constant.
3. We fix the **random seed** to ensure reproducibility of the sensitivity analyses.
4. The core function of the `lhs` package, `lhs::randomLHS()`, is only able to perform standard uniform sampling. Two parameters must be provided: the number of observations to derive (here, 5000), and the dimension of the random vector (here, $9$ parameters are considered).
5. How to generate the uniform distribution for a given parameter $X_j \sim \mathcal{U}(0.8 \times X_j^0, 1.2 \times  X_j^0)$, with $X_j^0$ the constant value estimated in a healthy individual, from the standard uniform distribution $U \sim \mathcal{U}(0, 1)$^[Note that we assume independence between the parameters, and that the *quantile call* is performed independently for each of the dimensions.]. By applying the **Inverse Transform Sampling Theorem** (see @tip-ITST), which implies applying the reciprocal of the CDF, in other words, the *quantile function* which is given in R by the `stats::qunif` function. Alternatively, we could have reproduced this result by applying the following affine transformation: `bootstrap_parameters[,i] * (param_ranges$UB[i] -param_ranges$LB[i]) +  param_ranges$LB[i]` on the observations simulated with a standard uniform. 
6. By applying the `quantile` function, and since the distribution is supposed evenly distributed on the sampling space, we expect the quantiles to be rather close from their theoretical values (for example, both the median and the mean should be close to the real constant parameter values).


::: {#tip-ITST .callout-tip collapse="true" title="The Inverse Transform Sampling Theorem"}

The **Inverse Transform Sampling Theorem** allows generating any probability distribution from the **standard uniform distribution**, $U \sim \mathcal{U}(0,1)$. Let $X$ be a continuous random variable with *cumulative distribution function* (CDF) $F(x)$ and $F^{-1}$ its reciprocal (the quantile function), then:

$$
X = F^{-1}(U)
$$

follows the same distribution as $X$.

---

In particular, for a uniform distribution $X \sim \mathcal{U}(a, b)$, $a$ and $b$ being respectively the lower and upper bounds, whose CDF is given by: $F(x) = \frac{x - a}{b - a}, \quad a \leq x \leq b$, then the *quantile* function is given by $F^{-1}(u) = a + u(b - a)$. 


In other terms, if $U$ follows the standard uniform distribution, then, by applying the **affine** transformation $X=a + (b - a) U$, $X$ will follow the $X \sim \mathcal{U}(a,b)$ distribution. 


Generally speaking, the **Inverse Transform Sampling** allows generating any probability distribution from a uniform $[0,1]$ applying this formula: $X = F^{-1}(U)$.

:::


`bootstrap_parameters` is a 5000 simulations times 9 parameters matrix storing all sampled parameter configurations. We displayed below only the first six lines of the simulated parameter distributions:

```{r bootstrap-parameters}
#| echo: false 
head(bootstrap_parameters)
```


### Compute steady states for the 5000 simulated parameter variations

We compute the corresponding steady-states for each of the $N=5000$ simulated parameter configurations. This task is computationally-intensive, but can be easily parallelised, since all bootstrap samples have been generated independently. We use to that end the standard `parallel` package.

```r
cl <- makeCluster(detectCores()) # <1>
clusterExport(cl, varlist = c("parameters_healthy", "bootstrap_parameters")) # <2>
cluster_environments <- clusterEvalQ(cl, { # <2>
  library(CoRC) # <2>
  library(dplyr, quietly = TRUE) # <2>
  healthy_model <- loadModel("./models/team_2016_final_model_lo2016.cps") # <2>
}) # <2>
sensitivity_outputs <- parLapply(cl = cl, # <3>
                    X = 1:n_samples, # <3>
                    f = function (i) { # <3>
                      sensitivity_model <- healthy_model |> saveModelToString() |>  loadModelFromString()  # <5>
                      setParameters(model = sensitivity_model, # <3>
                                    key = parameters_healthy$key, # <3>
                                    value = bootstrap_parameters[i,])  # <3>
                      sensitivity_model_steady_state <- runSteadyState(  # <3>
                        model = sensitivity_model # <3>
                      )$species # <3>
                      
                      SS_concentrations <- setNames(sensitivity_model_steady_state$concentration,  # <3>
                                                    sensitivity_model_steady_state$name) |>        # <3>
                        as.list() |> as.data.frame() |>                                            # <3>
                        dplyr::bind_cols(bootstrap_parameters[i, ,drop=FALSE] |>                   # <3>
                                           as.data.frame())                                        # <3>
                      return(SS_concentrations)                                                    # <3>
                    }
                    , chunk.size = 50) |>                                                         # <4>
                    dplyr::bind_rows()  # <6>

stopCluster(cl)  # <7>
```
1. Use `detectCores()` to determine the number of available CPU cores; conditionned to that number (usually 8 or 16 on recent personal laptop configurations); use `makeCluster()` to create the corresponding cluster^[With 8 cores, for instance, 8 computations can be performed in parallel. It might be advised to free completely a CPU to avoid computer freezing].
2. Use `clusterExport()` to send current session variables to workers, and `clusterEvalQ()` to load required libraries and/or execute code prior to the parallelised computations. Both functions ensure that the cluster environement has the required modules to start the computation. 
3. The Parallel Processing `parLapply()` is the counterpart of the sequential `lapply()` function. The first argument `cl` is the cluster definition. For each parameter configuration among the $N=5000$, we modifify the original ODE model to reflect the LHS parameter configuration, then, compute the modified steady-states associated with that change of parameter configuration. Finally, we return for each parameter configuration a dataset with the modified 15 steady-state concentrations and the corresponding 9 parameter configurations. 
4. The `chunk.size` option controls how data is divided among workers Once computed, another batch of 50 parameter configurations is processed).
5. This line is required, as `CoRC` uses copy by *reference* rather than by *value*. If we had just assigned the loaded healthy ODE model to the modified sensitivity model, `sensitivity_model <- healthy_model`, every modification on `sensitivity_model` would also be reported on `healthy_model`^[Another way to visualise the **copy by reference** approach: simply copy the ODE model in another variable `healthy_model_copy <- healthy_model` in the R canonical way. Compare the storage addresses: `lobstr::obj_addr(healthy_model)` and `lobstr::obj_addr(healthy_model_copy)`; you will note that they are identical].
6. The output of `parLapply` is a list of `data.frame`, function `dplyr::bind_rows()` concatenates them in a single large dataset.
7. Stop the cluster using `stopCluster()` to free computational resources when no longer need.

### Compute partial correlation scores

To compute the correlations between variations of the constant parameters, and the concentrations of the species at steady state, @lo2016po chose to apply a *rank-transformation* first, which guarantees increased robustness to outliers, and reduces *scaling issues* associated with significant variations of orders of magnitude between the parameters.

```{r rank-transformation}
# reshape and rank-transform sensitivity outputs
sensitivity_outputs <- readr::read_csv("../../results/sensitivity_outputs.csv") |>
  dplyr::mutate(across(where(is.numeric), rank)) 
```


Then, instead on relying on standard Pearson correlation which may yield spurious relationships between outcomes and parameter variations, @lo2016po computes the **Partial Correlation** score between the rank-transformed variations of species of interest, and the simulated parameter distributions (see @imp-prcc for details). 

The `pcor::ppcor` function returns both:

- `estimate`, the symmetric matrix of the partial correlation coefficient between two variables (included between -1 and 1, the sign providing the type of correlation, and the absolute value the magnitude or strength of the direct relationship)
- `p.value`, a matrix of the $p-$ values for each of the tests. 
- As in @lo2016po, the threshold for *significance* was set to $0.01$.

The resulting PRCC scores between T1 and T2 immune subpopulations with the 9 parameters tested for sensitivity are reported in @tbl-pcor-computation:

```{r}
#| label: tbl-pcor-computation
#| tbl-cap: PRCC values of the 9 parameters benchmarked for **robustness** against steady-state concentrations of $T_1$ and $T_2$ immune cell subtypes. Reproduction of Table 5 in @lo2016po. 

prcc_sensitivity <- pcor(sensitivity_outputs)
prcc_estimate <- prcc_sensitivity$estimate |> 
  tibble::as_tibble(mat, rownames = "species") |> 
  dplyr::select (species, T1, T2) |> 
  filter(!species %in% c("T1", "T2")) |> 
  # Join on the same dataset pvalues and PRCC estimations.
  inner_join(prcc_sensitivity$p.value |> 
               tibble::as_tibble(mat, rownames = "species") |> 
               dplyr::select (species, T1, T2) |> 
               filter(!species %in% c("T1", "T2")), by = "species",
             suffix = c("", ".pval")) |> 
  mutate(T1 = if_else(T1.pval <=0.01, sprintf("%.3f*", T1), sprintf("%.3f", T1)),
         T2 = if_else(T2.pval <=0.01, sprintf("%.3f*", T2), sprintf("%.3f", T2))) |> 
  dplyr::select(-T1.pval, -T2.pval)

### Format PRCC table
flextable(prcc_estimate) |> 
  add_footer_row(values = "* denotes significant PRCC with p-value below 0.01.", 
                 colwidths = 3) |> 
  compose(j = "species",
          value = as_paragraph(as_equation(c("\\sigma_{M_{\\alpha}}", "\\sigma_{M_{10}}",
                                             "\\sigma_{12}", "\\sigma_{2}", "\\sigma_{4}",
                                             "\\sigma_{21}", "\\sigma_{6}", 
                                             "\\sigma_{\\beta}", "\\sigma_{10}")))) |> 
  set_header_labels(values = list(
    species = "", T1 = "PRCC for T1", T2 = "PRCC for T2")) |> 
  align(align = "center", part = "all") |> 
  set_caption("Table 5: PRCC values for key parameters playing on T1 and T2 variations.") 
```


::: {#imp-prcc .callout-important collapse="true" title="Pearson Correlation and Partial Correlation"}

The standard **Pearson correlation** coefficient between two variables $X$ and $Y$ is given by @eq-pearson-correlation:  

$$
\rho_{X,Y} = \frac{\text{Cov}(X, Y)}{\sigma_X \sigma_Y}
$${#eq-pearson-correlation}

where:  
- $\text{Cov}(X, Y) = \mathbb{E}[(X - \mu_X)(Y - \mu_Y)]$ is the covariance.  
- $\sigma_X = \sqrt{\text{Var}(X)}$ and $\sigma_Y = \sqrt{\text{Var}(Y)}$ are the variable standard deviations. 

It measures the **linear relationship** between $X$ and $Y$, but does not control the effect of potential colliders or confounding variables.

---

On the other hand, the *Partial Correlation* measures the relationship between two variables while **controlling for** the effect of a the remaining set of variables. In a multi-dimensional framework, it's given by @eq-partial-correlation:

$$
\rho_{X,Y \mid \mathbf{Z}} = -\frac{\Omega_{X,Y}}{\sqrt{\Omega_{X,X} \Omega_{Y,Y}}}
$${#eq-partial-correlation}

where $\mathbf{\Omega} = \mathbf{\Sigma}^{-1}$ is the **precision matrix** (inverse of the covariance matrix).  

---

Compared to standard correlations, it removes indirect Relationships triggered by another set of confusing **variables** $\mathbf{Z}$, thus isolates the **direct** relationship between $X$ and $Y$, and enables more straightforward interpretation by revealing truly causal relationships. 
For example, in @fig-prcc-illustration, using simply the Pearson correlation, we would have certainly observed a significant correlation between variables $A$ and $C$. On the other hand, with partial correlation, the spurious connection between $A$ and $C$ due to the confusing effect of $B$ would certainly has been faded out. 

```{dot}
//| label: fig-prcc-illustration
//| fig-cap: "Simple example of **chain rule** associations."
digraph G {
    layout=neato
    A -> B;
    B -> C;
}
```

:::

### Plot rank-transformed datasets

```{r format-sensitivity-ggplot}
sensitivity_outputs_formatted <- sensitivity_outputs |>
  # tidy format: one column storing PRCC values for the x-axis, one for the y-axis, and another for the facets
  dplyr::select (c("T1", "T2", "sb", "s10", "s12")) |> 
  tidyr::pivot_longer(cols = c("sb", "s10", "s12"), names_to = "parameters", values_to = "x") |>
  tidyr::pivot_longer(cols = c("T1", "T2"), names_to = "species", values_to = "y") |> 
  # provide the same order of facets for direct comparison
  dplyr::mutate(parameters = factor(parameters, 
                                    levels = c("sb", "s10", "s12"),
                                    labels = c("sigma[b]", "sigma[10]", "sigma[12]"),
                                    ordered = TRUE),
                species = factor(species, 
                                 levels = c("T1", "T2"),
                                 labels = c("T[1]","T[2]")))

# Subtitles customised for each facet, including the p-value.
prcc_values <- prcc_sensitivity$estimate
prcc_sensitivity_annot <- data.frame(x=n_samples/2, y=Inf, 
                                     lab=factor(paste("PRCC =", c(prcc_values["T1", "sb"],
                                                           prcc_values["T1", "s10"],
                                                           prcc_values["T1", "s12"],
                                                           prcc_values["T2", "sb"],
                                                           prcc_values["T2", "s10"],
                                                           prcc_values["T2", "s12"]) |> 
                                                 sprintf(fmt = '%.3f')), ordered = TRUE), 
                                     parameters=factor(rep(c("sigma[b]", "sigma[10]", "sigma[12]"), 2), 
                                                       ordered = TRUE),
                                     species= factor(c(rep("T[1]", 3), rep("T[2]", 3)), 
                                                     ordered = TRUE))
```

We generate a scatter plot after applying rank-transformation of $T_1$ and $T_2$ immune cells subsets, against 3 parameters playing a critical role on the viability of cell pools, namely $\sigma_b$, $\sigma_{10}$ and $\sigma_{12}$, the first two controlling the activation rate of Tregs, and the induction of $T_1$ by the global pool of macrophages (including both $M1$ and $M2$ types)^[Note that we don't round the PRCC values, but only format them to display 3 significant digits using the native C `sprintf()`: difference between both approaches is notably crucial when saving datasets to avoid loss of numerical precision].

```{r}
#| label: fig-sensitivity-plots
#| fig-cap: Scatter plots of *rank-transformed* $T_1$ and $T_2$ against 3 parameters enumerated in @tbl-pcor-computation, exhbiting both strong partial correlation ($|\text{PRCC}|>0.5$ and $p-$value $<0.01$). Reproduction of [@lo2016po, pp. 11], Fig.2. 

SS_plots <- ggplot(sensitivity_outputs_formatted, aes(y = y, x=x)) +
  geom_point(size = 0.1, col = "blue", shape = 20) +
  # The labeller trick enables to use Latex notations for the subtitles of facets.
  facet_grid(species ~ parameters, 
             labeller = labeller(.rows = label_parsed, .cols = label_parsed)) +
  geom_label(aes(x, y, label=lab),
            data=prcc_sensitivity_annot, vjust=0.8, size = 3) +
  xlab("Cell Species") + ylab("Free and Variable Parameters") +
  # Guarantees minimal plot background that would hinder visibility of the output.
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.text.y = element_text(angle = 0)) 
SS_plots
```


One of the major advantages of the `ggplot2` package lies in its seamless ability to save the graphical object in a variety of formats, including `pdf` and `png`, as illustrated in @lst-ggplot-save: 

```{r ggplot2-illustration}
#| lst-label: lst-ggplot-save
#| lst-cap: Code snippet not run, used for illustration example. 
#| eval: false

# output is pdf, with a high resolution close to retina visual discrimination ability
ggsave(filename = "../../results/Fig2_PRCC.pdf", 
       plot = SS_plots, dpi = 600, width = 20, height = 10,
       units = "cm")
```
